---
title: "The Visual Raster Cheat Sheet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Visual Raster Cheat Sheet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = c("png"),
  fig.height = 3, 
  fig.width = 9, 
  comment = "#>"
)
```

#### Etienne B. Racine (etiennebr@gmail.com)
`r Sys.Date()` (R version 
`r paste0(version[["major"]], ".", version[["minor"]])`
and raster package version 
`r packageVersion("raster")`)

Source code and package available at https://github.com/etiennebr/visualraster. 
Suggestions, contributions, feature request, and bug report are appreciated and help to keep that project alive.

***

```{r echo = FALSE, cache = FALSE}
library("knitr")
library("xtable")

# modify stylesheet
#options(markdown.HTML.stylesheet = system.file('misc', 'vr_vignette.css', package='visualraster'))
# xtable params
options(xtable.type = "html")
options(xtable.include.rownames = FALSE)
```


```{r echo = FALSE, cache = FALSE, message = FALSE}
library(raster)
library(ggplot2)
library(patchwork)
library(RColorBrewer)
library("visualraster")

# define palette of colors
pal = list (
  qualitative = brewer.pal(11, "Set3"),
  circular = c("#0088FF", "#ffffff", "#F51D1D", "#400038", "#0088FF")
)
names(pal$qualitative) = 1:length(pal$qualitative)

proj_random_tm = "+proj=tmerc +lat_0=0 +lon_0=0 +x_0=0 +y_0=0 +units=m"

x <- as.vector(pal$qualitative)
cols <- c("0" = x[1], "1" = x[2], "2"= x[3], "3" = x[4], "4" = x[5], "5" = x[6],
          "6" = x[7], "7" = x[8], "8" = x[9], "9" = x[10])


theme_set(theme_void())

lh <- function(r, text = TRUE) {
  p <- 
    ggplot() + 
    vr_geom_raster(r)
  if(text) {
    p <- p + vr_geom_text(r) 
    }
  p <- 
    p +
    scale_colour_manual(values = cols,
                        aesthetics = c("colour", "fill"),
                        guide = "none")
  return(p)
}
# make randomness repeatable
set.seed(101)
```


# I/O
Input and output operations, load or write data

## raster
```{r}
r = raster(matrix(sample(1:9, 100, replace = TRUE), 10, 10))
```

```{r echo = FALSE}
p0 <- 
  lh(r) +
  labs(subtitle = "r")
p0
```

## write
```{r eval = FALSE}
writeRaster(r, "raster.tif")
```

## read
```{r eval = FALSE}
r = raster("raster.tif")
```

# Local
Local analysis consider cells independently

## Access

```{r echo = FALSE}
# generate point
pt = data.frame(x=0.22, y=0.72)
gpt = vr_geom_point(pt)

# generate line
stream = data.frame(
  x = cumsum(runif(7, max = 0.2)),
  y = cumsum(runif(7, max = 0.2))
)

ln = SpatialLines(list(Lines(list(Line(stream)), "1")) )
gln = vr_geom_line(stream)

# generate polygon
poly = SpatialPolygons(list(Polygons(list(Polygon(
  rbind(c(0.32, 0.74), 
        c(0.62, 0.84), 
        c(0.72, 0.25), 
        c(0.44, 0.25), 
        c(0.32, 0.74)))), 1)))
cpp  = fortify(poly)
gpy = vr_geom_polygon(cpp)
```

```{r echo = FALSE}
p1 <- 
  lh(r, FALSE) +
  #vr_geom_raster(r) + 
  vr_geom_text(r, label = 1:100, family = "mono", size = 3) +
  labs(subtitle = "Direct index `r[13]`")
```

```{r echo = FALSE}
grd = expand.grid(col = 1:10, lin = 1:10)
ref = with(grd, paste0(lin, ",", col))
p2 <- 
  lh(r) +
  vr_geom_raster(r) +
  vr_geom_text(r, family = "mono", size = 3, label = ref) +
  labs(subtitle = "Using line and columns `r[2, 3]`")
p0 + p1 + p2
```

```{r}
r[13]
r[2, 3]
r[2,  ]
r[ , 3]
#r[2:3, , drop = FALSE] |> lh()
```


Add vectors

```{r echo = FALSE}
# plot geometries
ggplot() + vr_geom_raster(r) + 
  vr_geom_text(r, family = "mono", size = 3, label = 1:100) +
  gln + gpt + gpy
```

### Cell from point
```{r}
cellFromXY(r, pt)
colFromX(r, pt$x)
rowFromY(r, pt$y)
fourCellsFromXY(r, as.matrix(pt))
```

### Cell from line
```{r}
cellFromLine(r, ln)
```

### Cell from polygon
```{r}
cellFromPolygon(r, poly)
```

## aggregate - disaggregate

```{r echo = FALSE}
r_agg = aggregate(r, 2, fun = mean)
r_dis = disaggregate(r, 2)
p1 <- lh(r_agg) + labs(subtitle = "aggregate(r, 2, fun = mean)")
p2 <- lh(r_dis) + labs(subtitle = "disaggregate(r, 2)")
p0 + p1 + p2
```

## cover

```{r echo = FALSE}
r_missing = raster(matrix(NA, 10, 10))
r_missing[41:60] = 0
r_covered = cover(r_missing, r)
p1 <- lh(r_missing) + labs(subtitle = "r_missing")
p2 <- lh(r_covered) + labs(subtitle = "cover(r_missing, r")
p0 + p1 + p2
```

## mask

```{r echo = FALSE}
r_masked  = mask(r, r_missing)
p2 <- lh(r_masked)  + labs(subtitle = "mask(r, r_missing)")
p0 + p1 + p2
```

## calc
```{r}
f = function(x) { x * 10 }
```

```{r echo = FALSE}
r_mul = calc(r, fun = f)
p1 <- 
  ggplot() + 
  vr_geom_raster(r_mul / 10) + 
  vr_geom_text(r_mul) +
  scale_colour_manual(values = cols,
                        aesthetics = c("colour", "fill"),
                        guide = "none") +
  labs(subtitle = "calc(r, fun = f)")
p0 + p1
```

## overlay
```{r, echo = FALSE}
r1 = init(r, function(x) sample(5:6, x, replace = TRUE))
r2 = init(r, function(x) sample(0:1, x, replace = TRUE))   # was 1:2
r3 = overlay(r1, r2, fun=sum)
p1 <- lh(r1) + labs(subtitle = "r1")
p2 <- lh(r2) + labs(subtitle = "r2")
p3 <- lh(r3) + labs(subtitle = "overlay(r1, r2, fun = sum)") 
p1 + p2 + p3
```

## distance
We need a raster containing origins to compute distances (so we need units)
```{r}
r_origin = raster(matrix(NA, 10, 10))
extent(r_origin) = extent(c(0, 10, 0, 10))
projection(r_origin) = "+proj=tmerc +lat_0=0 +lon_0=0 +x_0=0 +y_0=0 +units=m"
r_origin[c(15, 16, 45)] = 0
```

```{r echo = FALSE}
p1 <- lh(r_origin, FALSE) + labs(subtitle = "r_origin")
r_dist = distance(r_origin)
p2 <- 
  ggplot() + 
  vr_geom_raster(r_dist) + 
  vr_geom_text(r_dist) + 
  labs(subtitle = "distance(r_origin)") +
  scale_colour_manual(values = cols,
                        aesthetics = c("colour", "fill"),
                        guide = "none")
p1 + p2
```

## reclassify
```{r}
m = cbind(
  from    = c(0, 2, 4), 
  to      = c(2, 4, 8), 
  becomes = c(1, 2, 3)
)
```

```{r echo = FALSE, results = 'asis'}
xtable(m, digits = 0)
```

```{r, echo = FALSE}
r_reclass = reclassify(r_dist, m)
p3 <- lh(r_reclass) + labs(subtitle = "reclassify(r_dist, m)")
p2 + p3
```

# Focal
Focal analysis consider a cell plus its direct neighbours in a contiguous and symetrical manner

## focal
```{r}
r_focal = focal(r, fun = mean, w = matrix(1,nrow=3,ncol=3))
```

```{r, echo = FALSE}
p1 <- lh(r_focal) +
  #ggplot() + 
  #vr_geom_raster_seq(r_focal) + 
  #vr_geom_text(r_focal) +
  labs(subtitle = "focal(r, fun = mean, w = matrix(1,nrow=3,ncol=3))")
p0 + p1
```

One can remove edge effect by ignoring NA values
```{r, echo = FALSE}
r_focal_nedge = focal(r, fun = mean, w = matrix(1,nrow=3,ncol=3), na.rm = T, pad = T)
# ggplot() + vr_geom_raster_seq(r_focal_nedge) + vr_geom_text(r_focal_nedge)
p1 <- lh(r_focal_nedge) + labs(subtitle = "focal(r, fun = mean, w = matrix(1,nrow=3,ncol=3), na.rm = T, pad = T)")
p0 + p1
```

# Zonal
Zonal analysis consider group of cells in an irregular, but conitguous (in space or value) manner.

## zonal

QUO VADIS?

```{r echo = FALSE}
m = cbind(
  from    = c(0, 2, 4), 
  to      = c(2, 4, 8), 
  becomes = c(1, 2, 3)
)
z = reclassify(r_dist, m)
extent(z) = extent(r)
```

```{r echo = FALSE}
p1 <- p0 + labs(subtitle = "r")
p2 <- 
  ggplot() + 
  vr_geom_raster_seq(r) + 
  vr_geom_text(z)
p1 + p2
```

```{r eval = FALSE}
zonal(r, z, mean)
```

```{r echo = FALSE, results='asis'}
zones = zonal(r, z, mean)
colnames(zones) = gsub("structure.*$", "mean", colnames(zones))
xtable(zones, digits = c(0, 0, 1))
```

## clump
```{r echo = FALSE}
r_bin = raster(matrix(round(runif(100, max = 0.69)), 10, 10))
r_clumps = clump(r_bin)
p1 <- lh(r_bin) + labs(subtitle = "r_bin")
p2 <- lh(r_clumps) + labs(subtitle = "clump(r_bin)")
p1 + p2
```

## boundaries

"Detect boundaries (edges). boundaries are cells that have more than one class in the 4 or 8 cells surrounding it, or, if classes=FALSE, cells with values and cells with NA."

```{r, echo = FALSE}
r01 = init(r, function(x) ifelse(runif(x) > 0.15, 1, NA))
p1 <- lh(r01) + labs(subtitle = "r")
p2 <- lh(boundaries(r01)) + labs(subtitle = "boundaries(r)")
p1 + p2
```

## extract
```{r echo = FALSE}
# plot geometries, this time with values
p0 + gpt + gln + gpy
```

```{r}
extract(r, pt)
extract(r, ln)
extract(r, poly)
```

## rasterize

```{r echo = FALSE}
r_pt   = rasterize(pt, r)
r_line = rasterize(ln, r)
r_poly = rasterize(poly, r)

pts = as.data.frame(rasterToPoints(r))
gpts = geom_point(data = pts, aes(x = x, y = y), 
                  size = 2, shape = 20, color = I("#00000055"))
p1 <- lh(r_pt, FALSE) + gpt + labs(subtitle = "rasterize(pt, r)")
p2 <- lh(r_line, FALSE) + gln + labs(subtitle = "rasterize(ln, r)")
p3 <- lh(r_poly, FALSE) + gpy + labs(subtitle = "rasterize(poly, r)")
p1 + p2 + p3
```

Note that polygon rasterization is by default looking at cell centroid overlap, 

# Statistical
Statistical operations summarize the raster information

## density
```{r, eval = FALSE}
density(r)
```

```{r echo = FALSE}
qplot(r[], geom = "density", fill = I("grey20"))
```

## histogram
```{r, eval = FALSE}
hist(r)
```

```{r echo = FALSE}
qplot(r[], geom = "histogram", fill = I("grey20"), binwidth = 1)
```

## Spatial autocorrelation
```{r eval = FALSE}
Moran(r)
Geary(r)
```

```{r, echo = FALSE}
label_me = function (x, dec = 2) {
  paste0(
    "Moran = ", round(Moran(x), dec), "\n", 
    "Geary = ", round(Geary(x), dec)
  )
}
autocor_plot = function(x) {
  ggplot() +
    list(vr_geom_raster_seq(x),
         vr_geom_text(x, size = 2),
         ggtitle(label_me(x)))
}

r_plain = init(r, function(x) 1)
r_plain[c(55, 46)] = c(4, 0)

r_split = r
r_split[1:50] = 1
r_split[51:100] = 2

the_rs = list(r, 
              r_dist, 
              r_plain, 
              r_split, 
              init(r, fun = function(x) rep(1:2, length = x)),
              init(r, fun = function(x) rep(1:3, length = x)),
              init(r, fun = function(x) rep(1:4, length = x)),
              init(r, fun = function(x) rep(1:5, length = x)),
              init(r, fun = function(x) rep(1:6, length = x))
)
```

```{r echo = FALSE, fig.height = 9}
res <- list()
counter <- 0
for(w in the_rs) {
  counter <- counter + 1
  res[[counter]] <- autocor_plot(w)
  
}
res[[1]] + res[[2]] + res[[3]] + res[[4]] + res[[5]] +
  res[[6]] + res[[7]] + res[[8]] + res[[9]]
```

# Geometric
## extent
```{r eval = FALSE}
extent(r)
```

```{r echo = FALSE}
p0 +
  geom_polygon(data = extent_as_poly(r), aes(x = x, y = y), fill = "#ffffffaa", size = 1, colour = "black", linetype = "dashed")
```

## crop
```{r eval = TRUE}
e = extent(0.23, 0.86, 0.22, 0.73)
```


```{r echo = FALSE}
p1 <- 
  lh(r, FALSE) +
  geom_polygon(data = extent_as_poly(e), aes(x = x, y = y), fill = "#ffffffaa", size = 1, colour = "black", linetype = "dashed") + gpts +
  labs(subtitle = "r")

r_crop = crop(r, e)
p2 <- 
  lh(r_crop, FALSE) +
#ggplot() + vr_geom_raster(crop(r, f)) + 
  geom_polygon(data = extent_as_poly(e), 
               aes(x = x, y = y), 
               fill = NA, size = 1, 
               colour = "black",
               linetype = "dashed") + 
  geom_polygon(data = extent_as_poly(r), 
               aes(x = x, y = y), fill = NA, size = NA) + gpts +
  labs(subtitle = "crop(r, e)")
p1 + p2
```

## intersect
```{r}
g = extent(0.43, 1.1, -0.1, 0.53)
```
`

```{r echo = FALSE}
p1 <-
  lh(r, FALSE) +
  geom_polygon(data = extent_as_poly(g), 
               aes(x = x, y = y), 
               fill = NA, 
               size = 1, colour = "#111111", linetype = "dashed") +
  labs(subtitle = "r")

p2 <- 
  ggplot() + 
  geom_polygon(data = extent_as_poly(extent(r)), aes(x = x, y = y), fill = NA, size = 1, colour = "#aaaaaaaa", linetype = "dashed") + 
  #coord_equal() +
  vr_geom_raster(intersect(r, g)) + 
  geom_polygon(data = extent_as_poly(g), aes(x = x, y = y), fill = NA, size = 1, colour = "black", linetype = "dashed") +
  scale_colour_manual(values = cols,
                        aesthetics = c("colour", "fill"),
                        guide = "none") +
  labs(subtitle = "intersect(r, g)")
p1 + p2
```

### cellsFromExtent
```{r}
cellsFromExtent(r, g)
```


## union
```{r eval = FALSE}
union(extent(r), g)
```

```{r echo = FALSE}
r_union = union(extent(r), g)

ggplot() + 
  geom_polygon(data = extent_as_poly(g), aes(x = x, y = y), fill = NA, size = 2, colour = "#aaaaaaaa", linetype = "dashed") +
  geom_polygon(data = extent_as_poly(r), aes(x = x, y = y), fill = NA, size = 2, colour = "#aaaaaaaa", linetype = "dashed") +
  geom_polygon(data = extent_as_poly(r_union), aes(x = x, y = y), fill = NA, size = 2, colour = "#111111", linetype = "dashed")  + coord_equal()
```


## flip
```{r eval = FALSE}
flip(r_split, "y")
```

```{r echo = FALSE}
p1 <- lh(r_split) + labs(subtitle = "r_spit")
p2 <- lh(flip(r_split, "y")) + labs(subtitle = 'flip(r_split, "y")')
p1 + p2
```

## extend
```{r echo = FALSE}
p1 <- lh(extend(r, 3)) + labs(subtitle = "extend(r, 3)")
p0 + p1
```

## mosaic

```{r echo = FALSE, fig.height = 6}
# prepare data
r1 = raster::shift(r, dx = 0.3, dy = 0.2)
r1[] = 1
r2 = raster::shift(r, dx = -0.6, dy = -0.6)
r2[] = 1
r_mos = mosaic(r, r1, r2, fun = sum, na.rm = TRUE)
r_mos[is.nan(r_mos)] = NA
# plot rasters seperately
mos_grid = geom_tile(data = as_data(r_mos), 
                     aes(x = x, y = y, fill = NA), color = "#aaaaaa")
p1 <- ggplot() + mos_grid + vr_geom_raster(r) + vr_geom_text(r, size = 2) +
  scale_colour_manual(values = cols,
                        aesthetics = c("colour", "fill"),
                        guide = "none") +
  labs(subtitle = "r")
p2 <- ggplot() + mos_grid + vr_geom_raster(r1) + vr_geom_text(r1, size = 2) +
  scale_colour_manual(values = cols,
                        aesthetics = c("colour", "fill"),
                        guide = "none") +
  labs(subtitle = "r1")
p3 <- ggplot() + mos_grid + vr_geom_raster(r2) + vr_geom_text(r2, size = 2) + 
  scale_colour_manual(values = cols,
                        aesthetics = c("colour", "fill"),
                        guide = "none") +
  labs(subtitle = "r2")
p4 <- ggplot() + vr_geom_raster(r_mos) + vr_geom_text(r_mos, size = 2) +
  scale_colour_manual(values = cols,
                        aesthetics = c("colour", "fill"),
                        guide = "none") +
  labs(subtitle = "(mosaic(r, r1, r2, fun = sum, na.rm = TRUE)")
p1 + p2 + p3 + p4
```


## rasterToPoints
```{r}
pts = rasterToPoints(r)
```

```{r echo = FALSE, message = FALSE}
pts = as.data.frame(pts)
pts$coords = with(pts, paste0(round(x * 100), ",", round(y * 100)))
ggplot() +
  vr_geom_raster(r) +
  vr_geom_raster(r, color = "#ffffff", fill = "#ffffffaa") +
  geom_point(data = pts, aes(x = x, y = y, fill = factor(layer), label = coords), size = 6, shape = 21, color = I("#050505")) + 
  geom_point(data = pts, aes(x = x, y = y, color = factor(layer), label = coords), alpha = 0.5, size = 6, color = I("#050505"), shape = "+") + 
  geom_text(data = pts, aes(x = x, y = y, color = factor(layer), label = coords), size = 2, hjust = 0, vjust = 0, color = "#111111", angle = 45) +
  coord_equal()
```

## rasterToPolygons
```{r echo = FALSE, message = FALSE}
pols = rasterToPolygons(r_clumps, dissolve = TRUE)
p1 <- lh(r_clumps) + labs(subtitle = "r")
pols_fortify = fortify(pols)
p2 <- 
  ggplot() + 
  vr_geom_raster(r_clumps, color = "#111111") +
  geom_polygon(data = pols_fortify, aes(x = long, y = lat, group = group, color = id), fill = NA, size = 2) +
  geom_polygon(data = pols_fortify, aes(x = long, y = lat, group = group), fill = NA, size = 2, color = I("#00000055")) +
  scale_colour_manual(values = cols,
                        aesthetics = c("colour", "fill"),
                        guide = "none") +
  labs(subtitle = "rasterToPolygons(r, dissolve = T)")
p1 + p2
```

# Terrain
```{r echo = FALSE}
v = raster(volcano)
extent(v) = extent(0, ncol(v),0, nrow(v))
projection(v) =  proj_random_tm
```

```{r echo = FALSE}
p1 <- ggplot() + vr_geom_raster_seq(v) + labs(subtitle = "v") +
  tidyterra::scale_fill_whitebox_c()
p1
```

## terrain
```{r}
r_terrain = terrain(v, opt = c("slope", "aspect", "tpi", "tri", "roughness", "flowdir"))
```


```{r echo = FALSE, fig.height= 6}
#for(i in names(r_terrain)) 
#  print(ggplot() + vr_geom_raster_seq(r_terrain[[i]]) + labs(subtitle = (i)))
res <- list()
counter <- 0
for(i in names(r_terrain)) {
  counter <- counter + 1
  res[[counter]] <- ggplot() + vr_geom_raster_seq(r_terrain[[i]]) + labs(subtitle = (i)) +
    tidyterra::scale_fill_whitebox_c()
}
res[[1]] + res[[2]] + res[[3]] + res[[4]] + res[[5]] + res[[6]]
```

Angular data can sometimes be better expressed using a circular palette. 
In the following figure, blue is North orientation, while South is red. 
Both colors reach black at West and white at East. 
The preceeding figure had some sharp edges
on North faces, when angle slightly changed from 360 to 0.

```{r echo = FALSE, message = FALSE}
p1 <- 
  ggplot() + vr_geom_raster_seq(trim(r_terrain[["aspect"]])) + ggtitle("aspect") + scale_fill_gradientn(colours = pal$circular, limits = c(0, 2 * pi)) 

p2 <- 
  ggplot() + vr_geom_raster_seq(power_to_angle(trim(r_terrain[["flowdir"]]))) + ggtitle("flowdir") + scale_fill_gradientn(colours = pal$circular, limits = c(0, 360)) 
p1 + p2
```


## hillShade
```{r}
r_hill = hillShade(r_terrain[["slope"]], r_terrain[["aspect"]])
```

```{r echo = FALSE}
ggplot() + vr_geom_raster_seq(r_hill) + ggtitle("Hillshade")
```

## interpolate
```{r, echo = FALSE}
require("fields")
survey = as.data.frame(v, xy = TRUE)
survey = survey[sample(1:nrow(survey), 50), ]
```

```{r, eval = TRUE}
tps = Tps(survey[, c("x", "y")], survey$layer)
r_interpolate = interpolate(v, tps)
```

```{r echo = FALSE, message = FALSE, fig.height = 6}
gsurvey = vr_geom_point(survey, size = 2)
p1 <- ggplot() + vr_geom_raster_seq(v) + ggtitle("Volcano") + gsurvey
p2 <- ggplot() + vr_geom_raster_seq(r_interpolate, limits = range(v[])) + ggtitle("Interpolation") + gsurvey
p3 <- ggplot() + vr_geom_raster_div(r_interpolate - v) + ggtitle("Differences") + gsurvey
p4 <- ggplot() + vr_geom_raster_seq(interpolate(v, tps, fun=predictSE)) + ggtitle("Standard Error") + gsurvey
p1 + p2 + p3 + p4
```

# layerize
```{r echo = FALSE}
p1 <- lh(r_clumps) + labs(subtitle = "r_clumps")
s = layerize(r_clumps)
dd = as_data(s, melt = TRUE)
levels(dd$variable) = gsub("X", "", levels(dd$variable))
dd$fill = factor(with(dd, as.integer(variable) * value))
dd$fill[dd$fill == 0] = NA
dd$fill = droplevels(dd$fill)
p2 <- 
  ggplot() +
  geom_tile(data = dd, 
            aes(x = x, y = y, fill = fill), 
            color = "#888888", size = 0.25) +
  facet_wrap(~variable) + 
  coord_equal() +
  labs(subtitle = "layerize(r_clumps)") +
  scale_colour_manual(values = cols,
                        aesthetics = c("colour", "fill"),
                        guide = "none")
p1 + p2
```
